---
title: "class 08: mini project"
author: "Brian Wells (PID: A69026838)"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
>Q1: How many different candy types are in this dataset?

The number of rows is the number of candies.

```{r}
nrow(candy)
```


>Q2: How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

>Q3: What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Caramel Apple Pop", ]$winpercent
```

>Q4:What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

>Q5: What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```


```{r}
##install.packages("skimr")
```

```{r}
library("skimr")
skim(candy)
```
>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The variable on a difference scale seems to be "win percent".

>Q7. What do you think a zero and one represent for the candy$chocolate column?

0 probably means that the candy does not have chocolate, 1 means it does.

>Q8. Plot a histogram of winpercent values.

```{r}
library(ggplot2)
ggplot(candy, aes(winpercent)) +
geom_histogram()
```

>Q9: Is the distribution of winpercent values symmetrical?

no, the distribution is not symmetrical

>Q10: Is the center of the distribution above or below 50%?

It is below 50%

>Q11: On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])
```
on average, chocolate is ranked higher.

>Q12: Is this difference statistically significant?

```{r}
choc <- (candy$winpercent[as.logical(candy$chocolate)])
fruit <- (candy$winpercent[as.logical(candy$fruity)])
t.test(choc,fruit)
```
The difference is statistically significant.

>Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>% 
  arrange(winpercent) %>%
  head(5)
```

Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% 
  arrange(winpercent) %>%
  tail(5)
```
Snickers, Kit Kat, Twix, Reese's miniatures, Reese's Peanut Butter Cup

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

>Q17. What is the worst ranked chocolate candy?

Sixlets

>Q18. What is the best ranked fruity candy?

Starburst



```{r}
#install.packages("ggrepel")
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 10)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$pricepercent, decreasing = T)
tail( candy[ord,c(11,12)], n=5 )
```
Tootsie Roll Midgies


>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
Nik L nip is the least popular.


>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy),pricepercent)) +
  geom_col()
```

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy),pricepercent)) +
  geom_point() + 
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") 
```
```{r}
#install.packages("corrplot")
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

chocolate and fruity.

>Q23. Similarly, what two variables are most positively correlated?

chocolate and winpercent



```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
```


```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
#install.packages("plotly")
library(plotly)
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus. This makes sense because fruity candy is almost always pluribus and hard. The other qualities (caramel, nuts, nougat, wafer) are often associated with chocolate candy.