---
title: "Class 14: RNAseq mini project"
author: "Brian Wells (PID: A69026838)"
format: pdf
---

The data for for hands-on session comes from GEO entry: GSE37704, which is associated with the following publication:

Trapnell C, Hendrickson DG, Sauvageau M, Goff L et al. "Differential analysis of gene regulation at transcript resolution with RNA-seq". Nat Biotechnol 2013 Jan;31(1):46-53. PMID: 23222703
#data import
```{r}
#| message: false
#| warning: false
library(DESeq2)
```

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peak
metadata = read.csv(metaFile, row.names=1)
head(colData)
```

```{r}
# Import countdata
countData = read.csv(countFile, row.names=1)
head(countData)
```
#data tidying
>Q. Complete the code below to remove the troublesome first column from countData

```{r}
# Note we need to remove the odd first $length col
counts <- as.matrix(countData[,-1])
head(counts)
```

```{r}
all(colnames(counts) == metadata$id)
```

>Q. How many genes do we have to start with?

```{r}
nrow(counts)
```
>Q. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns).

```{r}
# Filter count data where you have 0 read count across all samples.
to.rm.ind <- rowSums(counts) == 0
counts <- counts[!to.rm.ind,]
nrow(counts)
```




#DESeq stuff

```{r}
#| message: false
#| warning: false
dds <-  DESeqDataSetFromMatrix(countData=counts,
                             colData=metadata,
                             design=~condition)
dds <-  DESeq(dds)
res <- results(dds)
head(res)
```
>Q. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.

```{r}
summary(res)
```
```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

>Q. Improve this plot by completing the below code, which adds color and axis labels

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```


#Add annotation data

>Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below.

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    key=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```





#Save results

>Q. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory.

```{r}
res = res[order(res$pvalue),]
write.csv(res, file="deseq_results.csv")
```



#Visualization

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

```{r}
library(pathview)
```

```{r}
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```

```{r}
# Look at the first few down (less) pathways
head(keggres$less)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

```{r}
# A different PDF based output of the same data
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids <- substr(keggrespathways, start=1, stop=8)
keggresids
```


```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```
>Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-reguled pathways?

Yes. 

```{r}
keggrespathways.down <- rownames(keggres$less)[1:5]
keggresids.down <- substr(keggrespathways.down, start=1, stop=8)
keggresids.down
```


```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids.down, species="hsa")
```






##Gene ontology
```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

##Reactome

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

>Q: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

Cell cycle, mitotic was the most significant. The first almost matches, its just less specific with KEGG (cell cycle). After this, it differs pretty significantly. This may be due to the fact that reactome has a much larger database than KEGG and can be far more specific with the pathways. The first several pathways on reactome are specific portions of the cell cycle group. Also we were looking at both "up" and "down" in reactome. With reactome, we were only looking at "down".


