---
title: "Class5: Data Visualization"
author: "Brian Wells (PID: A69026838)"
format: pdf
---

## Using GGPLOT

To use ggplot2 we first need to install it on our computers. To do this we will use the function `install.packages()`.

Before I use any package functions, I have to laod them up with a `library()` call, like so:
```{r}
# install.packages("ggplot2")
library(ggplot2)
ggplot(cars)
```
```{r}
head(cars)
```
There's always the "base R" graphics system, i.e. `plot()`

```{r}
plot(cars)
```

To use ggplot, I need to spell out at least 3 things:
- data (the stuff I want to plot as a data.frame)
- aesthetics (aes() values - how the data map to the plot)
- geoms (how I want things drawn)

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```


```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth()
```

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE)
```

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  labs(title="Speed and Stopping Distances of Cars",
       x="Speed (MPH)", 
       y="Stopping Distance (ft)",
       caption="Dataset: 'cars'") +
  geom_smooth(method="lm", se=FALSE)+
  theme_bw()
```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

to get how many genes there are (how many rows)
```{r}
nrow(genes)
```

```{r}
colnames(genes)
ncol(genes)
```

```{r}
table(genes$State)
```

```{r}
round(table(genes$State)/nrow(genes)*100,2)
```

```{r}
p <- ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State)+
  geom_point()
p + scale_colour_manual(values=c("blue","grey","red"))+
  labs(title="Gene Expresion Changes Upon Drug Treatment",
       x="Control (no drug) ",
       y="Drug Treatment")
```



```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"
gapminder <- read.delim(url)
#install.packages("dplyr")  ## un-comment to install if needed
library(dplyr)
gapminder_2007 <- gapminder %>% filter(year==2007)
ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp) +
  geom_point()
ggplot(gapminder_2007)+
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point (alpha=0.5)
```

```{r}
gapminder_2007 <- gapminder %>% filter(year==2007)
ggplot(gapminder_2007)+
  aes(x=gdpPercap, y=lifeExp, color=pop) +
  geom_point (alpha=0.8)
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)

ggplot(gapminder_1957) + 
  aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop) +
  geom_point(alpha=0.7) + 
  scale_size_area(max_size = 15) 
```
```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)

ggplot(gapminder_1957) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.7) + 
  scale_size_area(max_size = 15) +
  facet_wrap(~year)
```


