---
title: "Class 10 (pt 2)"
author: "Brian Wells (PID: A69026838)"
format: pdf
---


We'll finish off lab 10

##Comparative structure analysis of Adenylate Kinase

We will use `bio3d` package for this analysis that starts with a single sequence .
We will also use the `msa` package from BioConductor.
First we need to install the `BiocManager` we install it same as we always do (`install.packages()`)

We use `BiocManager::install()` to install any other BioConductor package we want - like `msa` in this case.

```{r}
library(bio3d)
```

First we will pick a sequence "1ake_A" and use `get.seq()`
```{r}
aa <- get.seq("1ake_A")
```

```{r}
aa
```


>Q13. How many amino acids are in this sequence, i.e. how long is this sequence? 

214

I want to seqarch for all related structures
```{r}
b <- blast.pdb(aa)
```
```{r}
hits <- plot(b)
```

```{r}
head(b$hit.tbl)
```

```{r}
hits$pdb.id
```

```{r}
save(hits, b, file="blast_results.Rds")
```

```{r}
load("blast_results.Rds")
```

Now we will download all the related structures from the database with `getpds()`






# Search and retrieve ADK structures

```{r}
hits$pdb.id <- c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A')

```

```{r}
# Download related PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=T, gzip=T)
```

#Align and superpose structures


```{r}
# Align releated PDBs
pdbs <- pdbaln(files, fit = T, exefile="msa")
```

```{r}
# Vector containing PDB codes for figure axis
ids <- basename.pdb(pdbs$id)

# Draw schematic alignment
#plot(pdbs, labels=ids)

```

#Annotate collected PDB structures

```{r}
anno <- pdb.annotate(ids)
unique(anno$source)
```

```{r}
anno
```
#Principal component analysis


```{r}
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```

```{r}
dim(pdbs$xyz)
```

```{r}
plot(pc.xray, 1:2)
```


```{r}
# Calculate RMSD
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)

plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```

#Optional: further viewing optimization

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```

```{r}
#Plotting results with ggplot2
library(ggplot2)
library(ggrepel)

df <- data.frame(PC1=pc.xray$z[,1], 
                 PC2=pc.xray$z[,2], 
                 col=as.factor(grps.rd),
                 ids=ids)

p <- ggplot(df) + 
  aes(PC1, PC2, col=col, label=ids) +
  geom_point(size=2) +
  geom_text_repel(max.overlaps = 20) +
  theme(legend.position = "none")
p
```


#Optional: Normal Analysis Mode

```{r}
# NMA of all structures
modes <- nma(pdbs)
```

```{r}
plot(modes, pdbs, col=grps.rd)
```

>Q14. What do you note about this plot? Are the black and colored lines similar or different? Where do you think they differ most and why?

Generally the peaks at certain residue numbers line up, but the magnitude (mainly between the black and colored lines) differ greatly. They differ most between residue ~125 - ~150. This could be t




